This research used resources of the National Energy Research Scientific
Computing Center, a DOE Office of Science User Facility supported by the Office
of Science of the U.S. Department of Energy under Contract No.
DE-AC02-05CH11231. 

The authors thank Doug Jacobsen, Woo-Sun Yang, Tina Declerck and Rebecca
Hartman-Baker for assistance with the large scale runs at NERSC. Edgar
Solomonik, Penporn Koanantakool and Evangelos Georganas provided helpful comments
and suggestions on tuning the MPI codes. We would like to acknowledge Craig
Tull, Ben Bowen and Michael Wehner for providing scientific data sets used
in the study. We are also grateful to the Daya Bay Collaboration for giving
us access to their experimental data.

This research is partially funded by DARPA Award Number HR0011-12-2-0016, the
Center for Future Architecture Research, a member of STARnet, a Semiconductor
Research Corporation program sponsored by MARCO and DARPA, and ASPIRE Lab
industrial sponsors and affiliates Intel, Google, Hewlett-Packard, Huawei, LGE,
NVIDIA, Oracle, and Samsung.  This work is supported by Cray, Inc., the Defense
Advanced Research Projects Agency XDATA program and DOE Office of Science
grants DOE DE-SC0010200 DE-SC0008700, DE-SC0008699. AD is supported by the
National Science Foundation Graduate Research Fellowship under Grant No. DGE
1106400. Any opinions, findings, conclusions, or recommendations in this paper
are solely those of the authors and does not necessarily reflect the position
or the policy of the sponsors.
